---
title: "GLAMMGoF Tutorial"
output: 
  rmarkdown::html_document:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{GLAMMGoF Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T,
  warning = F,
  comment = NA,
  message = F,
  fig.width = 7, 
  fig.height = 4
)
```

###### What does GLAMMGoF do?

`GLAMMGoF` is an R package for resampling-based predictive validation of generalized linear and generalized additive models. The two main functions in this package, `BIAS_PRECISION()` and `BRIER_AUC()`, quantify in-sample and out-of-sample predictive performance by repeatedly refitting a model structure to resampled training data and evaluating predictions on held-out testing data. These functions support model objects from a variety of R packages, including `glmmTMB()` from `glmmTMB`, `glmer()` and `glmer.nb()` from `lme4`, `glm.nb()` from `MASS`, `gam()` from `mgcv`, and `glm()` and `lm()` from the stats package. Optional residual diagnostics can also be performed using the `DHARMa` package.

-   `BIAS_PRECISION()` is used for models with continuous or integer response variables (e.g., Gaussian, Poisson, negative binomial, Tweedie, or Gamma distributions) and returns relative root mean squared error (RRMSE), relative mean absolute error (RMAE), relative median absolute error (RMedAE), and relative bias (RBIAS).
-   `BRIER_AUC()` is used for models with binary response variables (i.e., logistic regression) and returns AUC, Brier score, and log loss.
-   Both functions accept models with or without random effects, but predictive performance is based on population-level predictions, meaning random effects are excluded during prediction.
-   Two resampling methods are available via the `method` argument:
    -   repeated random holdout (`method = "holdout"`, the default), which splits data into fixed training and test sets across replicates and is intuitive and widely used; and
    -   bootstrap resampling with out-of-bag evaluation (`method = "bootstrap"`), which samples with replacement and tends to make more efficient use of the available data, particularly for smaller datasets.

**Important note:** The functions in this package take an existing model fitted to the full dataset and assess the predictive performance of that model structure through repeated resampling. In each replicate, the model is refit to a resampled training dataset and evaluated on separate testing data. This workflow is intentional: fitting the final model to all available data maximizes information for inference, while resampling-based predictive validation provides a practical way to assess generalization without permanently withholding data. This approach is especially useful in ecological applications where sample sizes are often limited. In general, `method = "holdout"` more closely reflects prediction to independent unseen observations and may be more intuitive for assessing generalization, whereas `method = "bootstrap"` makes more efficient use of the available data by evaluating performance on out-of-bag observations and may provide more stable estimates of predictive performance when sample sizes are small.

###### Getting started

You've either downloaded and installed `GLAMMGoF` from R-universe, GitHub, DevOps, or a local zipped package file. Once installed, you can access the vignette by running `browseVignettes("GLAMMGoF")`. Note that `ggplot2` is a dependency, and functions from numerous other R packages are imported and used as well.

```{r, warning=FALSE, message = FALSE, comment=NA}
library(GLAMMGoF)
```

```{r, warning=FALSE, message = FALSE, comment=NA, echo=FALSE}
library(dplyr)
library(tidyr)
```

###### How do I use GLAMMGoF?

There are only two functions in this package, and which one you use depends on the type of model being assessed:

`BIAS_PRECISION()`: used for models with continuous or integer response variables and estimates four predictive performance metrics using repeated resampling: `RRMSE`, `RMAE`, `RMedAE`, and `RBIAS`. For all four metrics, values closer to 0 indicate better predictive performance (see `?BIAS_PRECISION` for details).

-   All three accuracy statistics (`RRMSE`, `RMAE`, and `RMedAE`) express prediction error as a percentage of the mean observed value, which makes them interpretable on a common scale regardless of the units or magnitude of the response. To recover the corresponding raw error in the original units of the response, multiply any of these values by the observed mean and divide by 100. For example, an RRMSE of 18% with a sample mean of 50 implies a root mean squared error of 9 in the original units.

-   `RRMSE (Relative Root Mean Squared Error)`: The square root of average squared prediction errors, expressed as a percentage of the mean observed value. Because errors are squared before averaging, RRMSE penalises large individual errors more heavily than small ones, making it sensitive to cases where the model produces occasional extreme mispredictions.

-   `RMAE (Relative Mean Absolute Error)`: The average absolute prediction error expressed as a percentage of the mean observed value. RMAE treats all errors proportionally to their size without the extra weight that squaring applies, and is often the most intuitive summary of typical prediction accuracy.

-   `RMedAE (Relative Median Absolute Error)`: The median absolute prediction error expressed as a percentage of the mean observed value. Because it is based on the median rather than the mean of the error distribution, RMedAE is the most robust of the three to outlying predictions and gives the best picture of accuracy for a typical observation when the error distribution is skewed.

-   Taken together, these three statistics tell a more complete story than any one alone. If `RRMSE` is notably larger than `RMAE`, that signals a handful of high-leverage mispredictions are inflating the squared-error average, which is a pattern `RMedAE` will often fail to reflect if the bulk of predictions are accurate. Conversely, close agreement among all three suggests errors are roughly symmetric and there are no extreme outliers driving the summary.

-   `RBIAS (Relative Bias)`: The mean signed prediction error expressed as a percentage of the mean observed value, where positive values indicate systematic under-prediction and negative values indicate systematic over-prediction. `RBIAS` is independent of the accuracy metrics above: a model can be nearly unbiased on average yet still produce large errors, or it can be highly biased while still ranking observations correctly. Reporting `RBIAS` alongside RRMSE and RMAE therefore distinguishes random prediction noise from systematic directional error.

`BRIER_AUC()`: used for models with binary response variables and estimates three predictive performance metrics using repeated resampling: `AUC`, `Brier score`, and `log loss`. `AUC` values closer to 1 (bounded between 0 and 1) indicate better classification performance, whereas `Brier score` (bounded between 0 and 1) and `log loss` (lower bound of 0, no upper bound) values closer to 0 indicate better probabilistic predictions (see `?BRIER_AUC` for details).

-   The three binary metrics each capture a different aspect of predictive performance and are most informative when interpreted together.

-   `AUC (Area Under the ROC Curve)`: The probability that a randomly chosen positive case receives a higher predicted probability than a randomly chosen negative case, ranging from 0.5 (no discrimination) to 1 (perfect discrimination). `AUC` is purely a measure of discrimination (the model's ability to rank cases by their probability of the outcome) and is insensitive to how well the predicted probabilities are calibrated in absolute terms.

-   `Brier Score`: The mean squared difference between predicted probabilities and observed binary outcomes (0/1), ranging from 0 (perfect) to 1 (worst). The `Brier score` jointly rewards good discrimination and good calibration, penalising confident but wrong predictions more heavily than uncertain ones. A model can have high `AUC` but a poor `Brier score`if its predicted probabilities are systematically too high or too low even while correctly ordering cases.

-   `Log Loss (Binary Cross-Entropy)`: The negative mean log-likelihood of the observed outcomes under the predicted probabilities, where lower values indicate better performance. Like the Brier score, log loss rewards calibration, but its logarithmic scale means it penalizes extreme, confident mispredictions far more aggressively. `Log loss` is therefore the most sensitive of the three to cases where the model assigns near-zero probability to an outcome that actually occurs.

-   Reporting all three together is recommended: AUC reflects whether the model correctly separates cases from non-cases; Brier score provides an overall summary of probability accuracy on a squared-error scale; and log loss highlights whether the model is making any dangerously overconfident errors that the Brier score might underweight.

Both functions assess in- and out-of-sample predictive performance for generalized linear and generalized additive models, with or without random effects, by repeatedly resampling data and evaluating predictive performance across multiple replicates.

Two resampling methods are available via the method argument. The default (`method = "holdout"`) uses repeated random holdout (Monte Carlo cross-validation), which splits the data into a training set (80% by default, controlled by `propTrain`) and a test set (the remaining 20%). The model is refit to the training data and performance is evaluated on both the training data (in-sample performance) and the withheld test data (out-of-sample performance). This method is intuitive and widely understood, but the fixed training/test split means the training set is always smaller than the full dataset, which can be a limitation for smaller datasets.

Alternatively, `method = "bootstrap"` uses bootstrap resampling with replacement, where each bootstrap training dataset contains the same number of observations as the original dataset, sampled with replacement. Out-of-sample performance is evaluated on the out-of-bag observations not selected in the bootstrap sample (approximately 36.8% of observations on average). This method makes more efficient use of the available data and tends to produce more stable estimates, particularly for smaller datasets, but is slightly less intuitive than the holdout approach.

For well-behaved models and reasonably sized datasets, both methods should produce similar results. Differences are most likely to emerge with small datasets, highly overdispersed data, or poorly specified models.

This entire process is repeated `nReps` times, and the resulting sampling distributions of the performance metrics are summarized. Because `RRMSE`, `RMAE`, `RMedAE`, `RBIAS`, `AUC`, `Brier score`, and `log loss` quantify different aspects of predictive performance, they should be interpreted as complementary rather than interchangeable

There are two example datasets, `countData` and `logitData`, with integer and binary response variables, respectively. Each dataset also includes a variety of continuous and categorical predictor variables. Additionally, there are 12 example model objects (6 beginning with `count` and 6 beginning with `logit`): `Model_GLM`, `Model_GLMM`, `Model_GLMM2`, `Model_GAM`, `Model_GAMM`, and `Model_GAMM2`. Models ending in `M` include only fixed effects, whereas models ending in `MM` and `MM2` include one and two random effects, respectively. All 12 example models were fitted using `glmmTMB`, but `GLAMMGoF` supports models fitted using a variety of compatible R packages

###### Using the `BIAS_PRECISION()` function

To use this function properly, you must:

1.  Be sure you're using it with the correct model type, which in this case is one with a continuous or integer response variable. Here, a variety of distributions (e.g., Gaussian, Poisson, negative binomial, Tweedie, and gamma) are supported, as are numerous R packages (`glmmTMB`, `lme4` (including `glmer.nb`), `MASS` (`glm.nb`), `mgcv` (`gam`), and `glm`/`lm` from the `stats` package).

    1.  Note that if you include a random effect in a gam from mgcv, that variable MUST be a factor or else mgcv will throw an error.

2.  The only other choices you need to make are:

    1.  The number of bootstrap or Monte Carlo replicates you want to (the default is 100 but this should be at least 500 or 1000).

    2.  For method = "holdout" only: The proportion of data you want to use for assessing in-sample predictive performance (the default is 80%).

    3.  Whether you want to use the DHARMa package's `simulateResiduals()` function to assess the model's residuals (the default is TRUE), and how many simulation replicates you want DHARMa to use (the default is 1000).

Now let's run it and see what happens...

```{r, warning=FALSE, comment=NA}
bias_precision <- BIAS_PRECISION(nReps = 100, testModel = countModel_GLMM, testData = countData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 123, method = "holdout")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
bias_precision$bias_precision_results <- bias_precision$bias_precision_results %>% mutate_at(vars(value), funs(round(., 2)))
```

Look at the first few fit statistics...

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bias_precision$bias_precision_results, 8))
```

 

And then look at a summary of the `nReps` values...

```{r, warning=FALSE, comment=NA, echo=FALSE}
bias_precision$bias_precision_summary <- bias_precision$bias_precision_summary %>% mutate_at(vars(mn, lwr95, upr95), funs(round(., 2)))
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bias_precision$bias_precision_summary, 8))
```

 

And now look at a plot of the summaries

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
bias_precision$bias_precision_hist
```

 

Now let's compare to the `bootstrap` method, which generally should be very similar for well-behaved models and reasonably sized data sets.

```{r, warning=FALSE, comment=NA}
bias_precision <- BIAS_PRECISION(nReps = 100, testModel = countModel_GLMM, testData = countData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 123, method = "bootstrap")
```

Summary of `bootstrap` method

```{r, warning=FALSE, comment=NA, echo=FALSE}
bias_precision$bias_precision_summary <- bias_precision$bias_precision_summary %>% mutate_at(vars(mn, lwr95, upr95), funs(round(., 2)))
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bias_precision$bias_precision_summary, 8))
```

And now look at a plot of the summaries

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
bias_precision$bias_precision_hist
```

###### Using the `BRIER_AUC()` function

To use this function properly, you must:

1.  Be sure you're using it with the correct model type, which in this case is only a logistic regression with a binary response variable. Additionally, logistic regression models fit using the R packages `glmmTMB`, `lme4`, and `glm`/`lm` from the `stats` package are all supported.

2.  The only other choices you need to make are:

    1.  For method = "holdout" only: The proportion of data you want to use for assessing in-sample predictive performance (the default is 80%)

    2.  Whether you want to use the DHARMa package's `simulateResiduals()` function to assess the model's residuals (the default is TRUE).

```{r, warning=FALSE, comment=NA}
br_auc <- BRIER_AUC(nReps = 100, testModel = logitModel_GLMM, testData = logitData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 123, method = "holdout")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
br_auc$brier_auc_results <- br_auc$brier_auc_results %>% mutate_at(vars(value), funs(round(., 2)))
```

Look at the first few fit statistics...

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(br_auc$brier_auc_results, 6))
```

 

```{r, warning=FALSE, comment=NA, echo=FALSE}
br_auc$brier_auc_summary <- br_auc$brier_auc_summary %>% mutate_at(vars(mn, lwr95, upr95), funs(round(., 2)))
```

 

And then look at a summary of the `nReps` values...

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(br_auc$brier_auc_summary, 6))
```

 

And now look at a plot of the summaries

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
br_auc$brier_auc_hist
```

 

Let's again compare to the `bootstrap` method, which generally should be very similar for well-behaved models and reasonably sized data sets.

```{r, warning=FALSE, comment=NA}
br_auc <- BRIER_AUC(nReps = 100, testModel = logitModel_GLMM, testData = logitData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 123, method = "bootstrap")
```

Summary of `bootstrap` method

```{r, warning=FALSE, comment=NA, echo=FALSE}
br_auc$brier_auc_summary <- br_auc$brier_auc_summary %>% mutate_at(vars(mn, lwr95, upr95), funs(round(., 2)))
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(br_auc$brier_auc_summary, 6))
```

And now look at a plot of the summaries

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
br_auc$brier_auc_hist
```

###### Error messages when using `BIAS_PRECISION()`

For obvious reasons, the code below will cause `BIAS_PRECISION` quit and throw an error message:

```{r, warning=FALSE, comment=NA, eval=FALSE}
BIAS_PRECISION(nReps = 100, testModel = logitModel_GLMM, testData = logitData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 123, method = "holdout")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
try(BIAS_PRECISION(nReps = 100, testModel = logitModel_GLMM, testData = logitData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 123, method = "holdout"))
```

###### Error messages when using `BRIER_AUC()`

For equally obvious reasons, the code below will cause `BRIER_AUC` quit and throw an error message:

```{r, warning=FALSE, comment=NA, eval=FALSE}
BRIER_AUC(nReps = 100, testModel = countModel_GLMM, testData = countData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 123, method = "holdout")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
try(BRIER_AUC(nReps = 100, testModel = countModel_GLMM, testData = countData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 123, method = "holdout"))
```
