---
title: "GLAMMGoF Tutorial"
output: 
  rmarkdown::html_document:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{GLAMMGoF Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T,
  warning = F,
  comment = NA,
  message = F,
  fig.width = 7, 
  fig.height = 4
)
```

###### What does GLAMMGoF do?

`GLAMMGoF` is an R package that uses resampling to assess goodness-of-fit and predictive performance of generalized linear and generalized additive models. The two main functions in this package, `BIAS_PRECISION()` and `BRIER_AUC()`, resample several performance statistics and support model objects from a variety of R packages, including `glmmTMB()` from `glmmTMB`, `glmer()` and `glmer.nb()` from `lme4`, `glm.nb()` from `MASS`, `gam()` from `mgcv`, and `glm()` and `lm()` from the `stats` package. This package also uses the R package `DHARMa` to assess goodness-of-fit based on scaled model residuals.

- `BIAS_PRECISION()` is used for models with continuous or integer response variables (e.g., Gaussian, Poisson, negative binomial, Tweedie, or gamma distributions).
- `BRIER_AUC()` is used for models with binary response variables (i.e., logistic regression).
- Both functions accept models with or without random effects, but predictive performance is based on population-level predictions (i.e., random effects are 
  ignored).
- Two resampling methods are available via the `method` argument: repeated random holdout (`method = "holdout"`, the default), which splits data into fixed training 
  and test sets across replicates and is intuitive and widely used; and bootstrap resampling with out-of-bag evaluation (`method = "bootstrap"`), which samples with 
  replacement and tends to make more efficient use of the available data, particularly for smaller datasets.

**Note:** The functions in this package take an existing model fit to all available data and assess the performance of that model structure by (1) refitting the model to a training subset and (2) evaluating in- and out-of-sample predictive performance. This approach is intentional: using all available data for model fitting and inference is preferable, particularly in ecological studies where sample sizes are often limited, and resampling-based performance assessment provides a practical alternative to withholding a test set at the outset. This philosophy is especially relevant when using `method = "bootstrap"`, where the full dataset is always used for training and out-of-bag observations serve as the test set.

###### Getting started

You've either downloaded and installed `GLAMMGoF` from R-universe, GitHub, DevOps, or a local zipped package file. Once installed, you can access the vignette by running `browseVignettes("GLAMMGoF")`. Note that `ggplot2` is a dependency, and functions from numerous other R packages are imported and used as well.

```{r, warning=FALSE, message = FALSE, comment=NA}
library(GLAMMGoF)
```

```{r, warning=FALSE, message = FALSE, comment=NA, echo=FALSE}
library(dplyr)
library(tidyr)
```

###### How do I use GLAMMGoF?

There are only two functions in this package, and which one you need to use depends on the type of model you're assessing:

-   `BIAS_PRECISION()` : used for models with continuous or integer response variables and bootstraps four performance statistics: RRMSE, RMAE, RMedAE, and RBIAS. For all four statistics, values closer to 0 indicate a better performing model (see `?BIAS_PRECISION` for details).

-   `BRIER_AUC()` : used for models with binary response variables and bootstraps two performance statistics, AUC and Brier scores. AUC scores closer to 1 and Brier scores closer to 0 indicate a better performing model (see `?BRIER_AUC` for details).

These functions assess in- and out-of-sample predictive performance for Generalized Linear and Additive Mixed Models with and without random effects by repeatedly resampling data and evaluating RRMSE, RMAE, RMedAE, and RBIAS (`BIAS_PRECISION`) or Brier scores and AUC (`BRIER_AUC`) statistics.

Two resampling methods are available via the `method` argument. The default (`method = "holdout"`) uses repeated random holdout (Monte Carlo cross-validation), which splits the data into a training set (80% by default, controlled by `propTrain`) and a test set (the remaining 20%). The model is refit to the training data and performance is evaluated on both the training data (in-sample performance) and the withheld test data (out-of-sample performance). This method is intuitive and widely understood, but the fixed training/test split means the training set is always smaller than the full dataset, which can be a limitation for smaller datasets.

Alternatively, `method = "bootstrap"` uses bootstrap resampling with replacement, where the training set is the same size as the original data and out-of-sample performance is evaluated on the out-of-bag observations not selected in the bootstrap sample (approximately 36.8% of observations on average, a consequence of sampling with replacement). This method makes more efficient use of the available data and tends to produce more stable estimates, particularly for smaller datasets, but is slightly less intuitive than the holdout approach.

For well-behaved models and reasonably sized datasets, both methods should produce similar results. Differences are most likely to emerge with small datasets, highly overdispersed data, or poorly specified models.

This entire process is repeated `nReps` times and the sampling distributions of the fit statistics are then summarized. RRMSE, RMAE, RMedAE, and RBIAS, and/or Brier scores and AUC are reported because they are complementary measures of predictive performance.

There are two example data sets, `countData` and `logitData` , with integer and binary response variables, respectively. Each data set has a variety of continuous and categorical predictor variable as well. Additionally, there are 12 example model objects (6 starting with `count` and 6 starting with `logit`: `Model_GLM`, `ModelGLMM`, `ModelGLMM2`, `ModelGAM`, `ModelGAMM`, and `ModelGAMM2`. Models ending in `M` include only fixed effects, whereas models ending in `MM` and `MM2` contain one and two random effects, respectively. All 12 example models were fitted using the R package `glmmTMB`, but a variety of R packages that fit Generalized Linear and Additive Mixed Models are also compatible with `GLAMMGoF`.

###### Using the `BIAS_PRECISION()` function

To use this function properly, you must:

1.  Be sure you're using it with the correct model type, which in this case is one with a continuous or integer response variable. Here, a variety of distributions (e.g., Gaussian, Poisson, negative binomial, Tweedie, and gamma) are supported, as are numerous R packages (`glmmTMB`, `lme4` (including `glmer.nb`), `MASS` (`glm.nb`), `mgcv` (`gam`), and `glm`/`lm` from the `stats` package).

    1.  Note that if you include a random effect in a gam from mgcv, that variable MUST be a factor or else mgcv will throw an error.

2.  The only other choices you need to make are:

    1.  The number of bootstrap or Monte Carlo replicates you want to (the default is 100 but this should be at least 500 or 1000).

    2.  For method = "holdout" only: The proportion of data you want to use for assessing in-sample predictive performance (the default is 80%).

    3.  Whether you want to use the DHARMa package's `simulateResiduals()` function to assess the model's residuals (the default is TRUE), and how many simulation replicates you want DHARMa to use (the default is 1000).

Now let's run it and see what happens...

```{r, warning=FALSE, comment=NA}
bias_precision <- BIAS_PRECISION(nReps = 10, testModel = countModel_GLMM, testData = countData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 100, seed = 42, method = "holdout")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
bias_precision$bias_precision_results <- bias_precision$bias_precision_results %>% mutate_at(vars(value), funs(round(., 2)))
```

Look at the first few fit statistics...

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bias_precision$bias_precision_results))
```

 

And then look at a summary of the `nReps` values...

```{r, warning=FALSE, comment=NA, echo=FALSE}
bias_precision$bias_precision_summary <- bias_precision$bias_precision_summary %>% mutate_at(vars(mn, lwr95, upr95), funs(round(., 2)))
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bias_precision$bias_precision_summary))
```

 

And now look at a plot of the summaries

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
bias_precision$bias_precision_hist
```

 

Now let's compare to the `bootstrap` method, which generally should be very similar for well-behaved models and reasonably sized data sets. 

```{r, warning=FALSE, comment=NA}
bias_precision <- BIAS_PRECISION(nReps = 100, testModel = countModel_GLMM, testData = countData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 42, method = "bootstrap")
```

Summary of `bootstrap` method

```{r, warning=FALSE, comment=NA, echo=FALSE}
bias_precision$bias_precision_summary <- bias_precision$bias_precision_summary %>% mutate_at(vars(mn, lwr95, upr95), funs(round(., 2)))
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bias_precision$bias_precision_summary))
```

And now look at a plot of the summaries

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
bias_precision$bias_precision_hist
```

###### Using the `BRIER_AUC()` function

To use this function properly, you must:

1.  Be sure you're using it with the correct model type, which in this case is only a logistic regression with a binary response variable. Additionally, logistic regression models fit using the R packages `glmmTMB`, `lme4`, and `glm`/`lm` from the `stats` package are all supported.

2.  The only other choices you need to make are:

    1.  For method = "holdout" only: The proportion of data you want to use for assessing in-sample predictive performance (the default is 80%)

    2.  Whether you want to use the DHARMa package's `simulateResiduals()` function to assess the model's residuals (the default is TRUE).

```{r, warning=FALSE, comment=NA}
br_auc <- BRIER_AUC(nReps = 100, testModel = logitModel_GLMM, testData = logitData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 42, method = "holdout")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
br_auc$brier_auc_results <- br_auc$brier_auc_results %>% mutate_at(vars(value), funs(round(., 2)))
```

Look at the first few fit statistics...

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(br_auc$brier_auc_results))
```

 

```{r, warning=FALSE, comment=NA, echo=FALSE}
br_auc$brier_auc_summary <- br_auc$brier_auc_summary %>% mutate_at(vars(mn, lwr95, upr95), funs(round(., 2)))
```

 

And then look at a summary of the `nReps` values...

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(br_auc$brier_auc_summary))
```

 

And now look at a plot of the summaries

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
br_auc$brier_auc_hist
```

 

Let's again compare to the `bootstrap` method, which generally should be very similar for well-behaved models and reasonably sized data sets. 

```{r, warning=FALSE, comment=NA}
br_auc <- BRIER_AUC(nReps = 100, testModel = logitModel_GLMM, testData = logitData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 42, method = "bootstrap")
```

Summary of `bootstrap` method

```{r, warning=FALSE, comment=NA, echo=FALSE}
br_auc$brier_auc_summary <- br_auc$brier_auc_summary %>% mutate_at(vars(mn, lwr95, upr95), funs(round(., 2)))
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(br_auc$brier_auc_summary))
```

And now look at a plot of the summaries

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
br_auc$brier_auc_hist
```

###### Error messages when using `BIAS_PRECISION()`

For obvious reasons, the code below will cause `BIAS_PRECISION` quit and throw an error message:

```{r, warning=FALSE, comment=NA, eval=FALSE}
BIAS_PRECISION(nReps = 100, testModel = logitModel_GLMM, testData = logitData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 42, method = "holdout")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
try(BIAS_PRECISION(nReps = 100, testModel = logitModel_GLMM, testData = logitData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 42, method = "holdout"))
```

###### Error messages when using `BRIER_AUC()`

For equally obvious reasons, the code below will cause `BRIER_AUC` quit and throw an error message:

```{r, warning=FALSE, comment=NA, eval=FALSE}
BRIER_AUC(nReps = 100, testModel = countModel_GLMM, testData = countData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 42, method = "holdout")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
try(BRIER_AUC(nReps = 100, testModel = countModel_GLMM, testData = countData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 42, method = "holdout"))
```
