---
title: "GLAMMGoF Tutorial"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: 
      collapsed: true
      smooth_scroll: true
    toc_depth: 4  
    code_overflow: wrap
    number_sections: false
vignette: >
  %\VignetteIndexEntry{GLAMMGoF Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T,
  warning = F,
  comment = NA,
  message = F,
  fig.width = 7, 
  fig.height = 4
)
```

### What does GLAMMGoF do?

`GLAMMGoF` is an R package for resampling-based predictive assessment of generalized linear and generalized additive models. The two main functions in the package, `bias_precision()` and `brier_auc()`, quantify in-sample and out-of-sample predictive performance by repeatedly refitting a model structure to resampled training data and evaluating predictions on held-out testing data. These functions support model objects from a variety of R packages, including `glmmTMB()` from `glmmTMB`, `lmer()`, `glmer()`, and `glmer.nb()` from `lme4`, `glm.nb()` from `MASS`, `gam()` from `mgcv`, and `glm()` and `lm()` from the stats package. Optional residual diagnostics can also be performed using the `DHARMa` package.

-   `bias_precision()` is intended for models with continuous or integer response variables (e.g., gamma, Gaussian, negative binomial, Poisson, or Tweedie distributions) and returns relative root mean squared error (RRMSE), relative mean absolute error (RMAE), relative median absolute error (RMedAE), and relative bias (RBIAS).

-   `brier_auc()` is intended for models with binary response variables (i.e., logistic regression) and returns AUC, Brier score, and log loss.

Both functions accept models with or without random effects. By default, predictive performance is based on population-level (marginal) predictions, meaning random effects are excluded during prediction (`re.form = ~0`). This behavior can be changed via the `conditional_predictions` argument (see "Marginal vs conditional predictions" below for details).

Two resampling methods are available via the method argument: repeated random holdout (method = "holdout", the default), which repeatedly partitions the data into training and testing subsets and is intuitive and widely used; and bootstrap resampling with out-of-bag evaluation (method = "bootstrap"), which samples with replacement and tends to make more efficient use of the available data, particularly for smaller data sets.

**Important note:** The functions in this package take an existing model fitted to the full data set and assess the predictive performance of that model structure through repeated resampling. In each of `nReps` replicates, the model is refit to a resampled training data set and evaluated on separate testing data. This workflow is intentional: fitting the final model to all available data maximizes information available for inference, while resampling-based predictive assessment provides a practical way to evaluate generalization without permanently withholding data. This approach is especially useful in ecological applications where sample sizes are often limited. In general, `method = "holdout"` more closely reflects prediction to independent unseen observations and may be more intuitive for assessing generalization, whereas `method = "bootstrap"` makes more efficient use of the available data by evaluating performance on out-of-bag observations and may provide more stable estimates of predictive performance when sample sizes are small.

### Getting started

You've either downloaded and installed `GLAMMGoF` from R-universe, GitHub, DevOps, or a local zipped package file. Once installed, you can access the vignette by running `browseVignettes("GLAMMGoF")`. Note that `ggplot2` is a dependency, and functions from numerous other R packages are imported and used as well.

```{r, warning=FALSE, message = FALSE, comment=NA}
library(GLAMMGoF)
```

```{r, warning=FALSE, message = FALSE, comment=NA, echo=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
```

### How is GLAMMGoF used?

There are only two functions in this package, and which one you use depends on the type of model being assessed:

`bias_precision()`: used for models with continuous or integer response variables and estimates four predictive performance metrics using repeated resampling: `RRMSE`, `RMAE`, `RMedAE`, and `RBIAS`. For all four metrics, values closer to 0 indicate better predictive performance (see `?bias_precision` for details). All three accuracy statistics (`RRMSE`, `RMAE`, and `RMedAE`) express prediction error as a percentage of the mean observed value, which places them on a common scale regardless of the units or magnitude of the response. To recover the corresponding raw error in the original units of the response, multiply any of these values by the observed mean and divide by 100. For example, an `RRMSE` of 18% with a sample mean of 50 implies a root mean squared error of 9 in the original units.

-   `RRMSE (Relative Root Mean Squared Error)`: The square root of average squared prediction errors, expressed as a percentage of the mean observed value. Because errors are squared before averaging, RRMSE penalizes large individual errors more heavily than small ones, making it sensitive to cases where the model produces occasional extreme predictions.

    ```         
    fit_cost_rrmse <- function(y, yhat) sqrt(mean((yhat - y)^2)) / mean(y) * 100
    ```

-   `RMAE (Relative Mean Absolute Error)`: The average absolute prediction error expressed as a percentage of the mean observed value. `RMAE` treats all errors proportionally to their size without the extra weight that squaring applies, and is often the most intuitive summary of typical prediction accuracy.

    ```         
    fit_cost_rmae  <- function(y, yhat) mean(abs(yhat - y)) / mean(y) * 100
    ```

-   `RMedAE (Relative Median Absolute Error)`: The median absolute prediction error expressed as a percentage of the mean observed value. Because it is based on the median rather than the mean of the error distribution, `RMedAE` is the most robust of the three to outlying predictions and gives the best picture of accuracy for a typical observation when the error distribution is skewed.

    ```         
    fit_cost_rmedae  <- function(y, yhat) median(abs(yhat - y)) / mean(y) * 100
    ```

-   Taken together, these three statistics tell a more complete story than any one alone. If `RRMSE` is notably larger than `RMAE`, that signals a handful of high-leverage predictions are inflating the squared-error average, which is a pattern `RMedAE` will often fail to reflect if the bulk of predictions are accurate. Conversely, close agreement among all three suggests errors are roughly symmetric and there are no extreme outliers driving the summary.

-   `RBIAS (Relative Bias)`: The mean signed prediction error expressed as a percentage of the mean observed value, where positive values indicate systematic over-prediction and negative values indicate systematic under-prediction. `RBIAS` is independent of the accuracy metrics above: a model can be nearly unbiased on average yet still produce large errors, or it can be highly biased while still ranking observations correctly. Reporting `RBIAS` alongside `RRMSE` and `RMAE` therefore distinguishes random prediction noise from systematic directional error.

    ```         
    fit_cost_rbias <- function(y, yhat) mean(yhat - y) / mean(y) * 100
    ```

`brier_auc()`: used for models with binary response variables and estimates three predictive performance metrics using repeated resampling: `AUC`, `Brier score`, and `log loss`. `AUC` values closer to 1 (bounded between 0 and 1) indicate better classification performance, whereas `Brier score` (bounded between 0 and 1) and `log loss` (lower bound of 0, no upper bound) values closer to 0 indicate better probabilistic predictions (see `?brier_auc` for details).

-   The three binary metrics each capture a different aspect of predictive performance and are most informative when interpreted together.

-   `AUC (Area Under the ROC Curve)`: The probability that a randomly chosen positive case receives a higher predicted probability than a randomly chosen negative case, ranging from 0 (no discrimination) to 1 (perfect discrimination), and where an `AUC` of 0.5 indicates that a model predicts no better than random chance (and `AUC` \<0.5 is worse than random). `AUC` is purely a measure of discrimination (the model's ability to rank cases by their probability of the outcome) and is insensitive to how well the predicted probabilities are calibrated in absolute terms. This function also returns the null model (i.e., intercept-only) `AUC` statistic, which is assumed to be 0.5, which allows for comparison to the summarized `AUC` metrics.

-   `Brier Score`: Across all observations, the mean squared difference between predicted probabilities and observed binary outcomes (0/1), ranging from 0 (perfect) to 1 (worst). The `Brier score` jointly rewards good discrimination and good calibration, penalizing confident but wrong predictions more heavily than uncertain ones. A model can have high `AUC` but a poor `Brier score` if its predicted probabilities are systematically too high or too low even while correctly ordering cases. This function also returns the null model (i.e., intercept-only) `Brier score` for reference (i.e., the best predictive performance achievable without any predictors), computed as `p_bar * (1 - p_bar)`, where `p_bar` is the observed event rate, which allows for comparison to the summarized Brier score metrics. Note that the null Brier score is based on the arithmetic mean of the response, which approximates but is not identical to plogis(intercept) from a fitted intercept-only logistic model; the difference is negligible in practice and avoids the overhead of model fitting.

-   `Log Loss (Binary Cross-Entropy)`: Across all observations, the mean negative log-likelihood of the observed outcomes under the predicted probabilities, where lower values indicate better performance. Like the `Brier score`, `log loss` rewards calibration, but its logarithmic scale means it penalizes extreme, confidently wrong predictions far more aggressively. `Log loss` is therefore the most sensitive of the three to cases where the model assigns near-zero probability to an outcome that actually occurs. This function also returns the null model (i.e., intercept-only) log loss for reference (i.e., the best predictive performance achievable without any predictors), computed as `-(p_bar * log(p_bar) + (1 - p_bar) * log(1 - p_bar))`, where `p_bar` is the observed event rate, which allows for comparison to the summarized log loss metrics. Note that the null log loss is based on the arithmetic mean of the response, which approximates but is not identical to plogis(intercept) from a fitted intercept-only logistic model; the difference is negligible in practice and avoids the overhead of model fitting. Log loss is computed internally as:

```         
    log_loss <- function(y, p, eps = 1e-15) {p <- pmax(pmin(p, 1 - eps), eps) -mean(y * log(p) + (1 - y) * log(1 - p))}
```

-   Reporting all three together is recommended: `AUC` reflects whether the model correctly separates cases from non-cases; `Brier score` provides an overall summary of probability accuracy on a squared-error scale; and `log loss` highlights whether the model is making any dangerously overconfident errors that the `Brier score` might underweight.

Both functions assess in- and out-of-sample predictive performance for generalized linear and generalized additive models, with or without random effects, by repeatedly resampling data and evaluating predictive performance across multiple replicates. Results are stored across all `nReps` replicates and reported as a summary table (mean and 95% confidence intervals) and a histogram of replicate values for each performance statistic, with a blue dotted vertical line indicating the mean across replicates. For `brier_auc()`, a red dashed vertical line is also displayed in the log loss panels, indicating the null (intercept-only) log loss baseline as a reference point for model comparison.

Two resampling methods are available via the method argument. The default (`method = "holdout"`) uses repeated random holdout (Monte Carlo cross-validation), which splits the data into a training set (80% by default, controlled by `propTrain`) and a test set (the remaining 20%). The model is refit to the training data and performance is evaluated on both the training data (in-sample performance) and the withheld test data (out-of-sample performance). This method is intuitive and widely understood, but the fixed training/test split means the training set is always smaller than the full data set, which can be a limitation for smaller data sets. Alternatively, `method = "bootstrap"` uses bootstrap resampling with replacement, where each bootstrap training data set contains the same number of observations as the original data set, sampled with replacement. Out-of-sample performance is evaluated on the out-of-bag observations not selected in the bootstrap sample (approximately 36.8% of observations on average). This method makes more efficient use of the available data and tends to produce more stable estimates, particularly for smaller data sets, but is slightly less intuitive than the holdout approach. For models without random effects and reasonably sized data sets, both methods should produce similar results. Differences can emerge for random-effect models, particularly when random-effect groups have few observations. See "A note about holdout vs bootstrap resampling methods when diagnosing and adjusting for Jensen's inequality bias" below for details. 

#### A note about zero-inflated models

For zero-heavy count or continuous data, the choice of performance metric matters. `RBIAS` in particular can behave unexpectedly when the response contains zeros or values near zero, because the relative error formulation divides by the observed value, producing unstable or extreme terms that can dominate the average and give a misleading impression of systematic underprediction. This is a known limitation of relative bias metrics rather than evidence of poor model fit (Hyndman and Koehler 2006), and is most likely to occur when zeros are common but the zero-generating process is not explicitly modeled.

-   In these situations, `RMedAE` or `RMAE` may provide more reliable summaries of predictive performance. `RMedAE` is robust to extreme errors because it is based on the median absolute error rather than the mean, and therefore better reflects predictive accuracy for a typical observation. `RMAE` can also be useful because absolute-error metrics weight errors linearly rather than quadratically, making them less sensitive to extreme observations than `RMSE`-based measures.

-   When combined with the `DHARMa` zero-inflation test (`testZI = TRUE`), these metrics provide a more complete assessment than any single measure alone. `RMedAE` or `RMAE` summarize predictive accuracy, while the ZI test evaluates whether the model correctly reproduces the observed frequency of zeros. A significant ZI test result with `ratioObsSim > 1` indicates that the model is underpredicting zeros, suggesting that a zero-inflated model family or additional sources of heterogeneity should be considered.

There are two example data sets, `countData` and `logitData`, with integer and binary response variables, respectively. Each data set also includes a variety of continuous and categorical predictor variables. Additionally, there are 12 example model objects (6 beginning with `count` and 6 beginning with `logit`): `Model_GLM`, `Model_GLMM`, `Model_GLMM2`, `Model_GAM`, `Model_GAMM`, and `Model_GAMM2`. Models ending in `M` include only fixed effects, whereas models ending in `MM` and `MM2` include one and two random effects, respectively. All 12 example models were fitted using `glmmTMB`, but `GLAMMGoF` supports models fitted using a variety of compatible `R` packages

### `bias_precision()`

#### Using the `bias_precision()` function

To use this function properly, you must:

1.  Be sure you're using it with the correct model type, which in this case is one with a continuous or integer response variable. Here, a variety of distributions (e.g., Gaussian, gamma, negative binomial, Poisson, or Tweedie) are supported, as are numerous R packages (`glmmTMB`, `lme4` (including `glmer`, `glmer.nb`, and `lmer`), `MASS` (`glm.nb`), `mgcv` (`gam`), and `glm`/`lm` from the `stats` package). The function will throw an informative error message if the wrong type of model is supplied. 

2.  The only other choices you need to make are:

    1.  The number of bootstrap or Monte Carlo replicates you want to use (the default is `nReps=100` but this should be at least 500 or 1000).

    2.  For `method = "holdout"` only: The proportion of data you want to use for assessing in-sample predictive performance (the default is 0.8).

    3.  Whether you want to use the `DHARMa` package's `simulateResiduals()` function to assess and plot the model's residuals (the default is `DHARMaPlot=TRUE`), whether you also want `DHARMa's` to test for zero inflation using the `testZeroInflation()` function (the default is `testZI=TRUE`), and how many simulation replicates you want `DHARMa` to use (the default is `DHARMaReps=1000`).

Now let's run it on countModel_GLMM and see what happens...

```{r, warning=FALSE, comment=NA}
bp <- bias_precision(nReps = 100, testModel = countModel_GLMM, testData = countData, propTrain = 0.8, DHARMaPlot = TRUE, testZI = TRUE, DHARMaReps = 1000, seed = 123, method = "holdout")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
bp$bias_precision_results <- bp$bias_precision_results %>% mutate(across(value, ~ round(., 2)))
```

Look at the first few fit statistics...

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bp$bias_precision_results, 8))
```

 

And then look at a summary of the `nReps` values...

```{r, warning=FALSE, comment=NA, echo=FALSE}
bp$bias_precision_summary <- bp$bias_precision_summary %>% mutate(across(c(mn, lwr95, upr95), ~ round(., 2)))
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bp$bias_precision_summary, 8))
```

 

And now look at a plot of the summaries...

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
bp$bias_precision_hist
```

 

Now let's compare to the `bootstrap` method for the same model, which generally should be very similar for well-behaved models and reasonably sized data sets (but see "A note about holdout vs bootstrap resampling methods when diagnosing and adjusting for Jensen's inequality bias" below for details)

```{r, warning=FALSE, comment=NA}
bp <- bias_precision(nReps = 100, testModel = countModel_GLMM, testData = countData, propTrain = 0.8, DHARMaPlot = TRUE, testZI = TRUE, DHARMaReps = 1000, seed = 123, method = "bootstrap")
```

Summary of `bootstrap` method

```{r, warning=FALSE, comment=NA, echo=FALSE}
bp$bias_precision_summary <- bp$bias_precision_summary %>% mutate(across(c(mn, lwr95, upr95), ~ round(., 2)))
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bp$bias_precision_summary, 8))
```

And now look at a plot of the summaries...

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
bp$bias_precision_hist
```

### `brier_auc()`

#### Using the `brier_auc()` function

To use this function properly, you must:

1.  Be sure you're using it with the correct model type, which in this case is only a logistic regression with a binary response variable. Additionally, logistic regression models fit using the R packages `glmmTMB`, `lme4`, and `glm`/`lm` from the `stats` package are all supported. The function will throw an informative error message if the wrong type of model is supplied. 

2.  The only other choices you need to make are:

    1.  The number of bootstrap or Monte Carlo replicates you want to use (the default `nReps = 100` but this should be at least 500 or 1000).

    2.  For `method = "holdout"` only: The proportion of data you want to use for assessing in-sample predictive performance (the default is `propTrain = 0.8`).

    3.  Whether you want to use the `DHARMa` package's `simulateResiduals()` function to assess and plot the model's residuals (the default is `DHARMaPlot=TRUE`) and how many simulation replicates you want `DHARMa` to use (the default is `DHARMaReps=1000`). There is no option to test for zero-inflation with this function as this is not strictly relevant to logistic regression models with binary responses.

```{r, warning=FALSE, comment=NA}
br_auc <- brier_auc(nReps = 100, testModel = logitModel_GLMM, testData = logitData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 123, method = "holdout")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
br_auc$brier_auc_results <- br_auc$brier_auc_results %>% mutate(across(value, ~ round(., 2)))
```

Look at the first few fit statistics...

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(br_auc$brier_auc_results, 9))
```

 

```{r, warning=FALSE, comment=NA, echo=FALSE}
br_auc$brier_auc_summary <- br_auc$brier_auc_summary %>% mutate(across(c(mn, lwr95, upr95), ~ round(., 2)))
```

 

And then look at a summary of the `nReps` values...

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(br_auc$brier_auc_summary, 9))
```

 

And now look at a plot of the summaries...

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
br_auc$brier_auc_hist
```

 

Let's again compare to the `bootstrap` method, which generally should be very similar for well-behaved models and reasonably sized data sets (but see "A note about holdout vs bootstrap resampling methods when diagnosing and adjusting for Jensen's inequality bias" below for details).

```{r, warning=FALSE, comment=NA}
br_auc <- brier_auc(nReps = 100, testModel = logitModel_GLMM, testData = logitData, propTrain = 0.8, DHARMaPlot = TRUE, DHARMaReps = 1000, seed = 123, method = "bootstrap")
```

Summary of `bootstrap` method

```{r, warning=FALSE, comment=NA, echo=FALSE}
br_auc$brier_auc_summary <- br_auc$brier_auc_summary %>% mutate(across(c(mn, lwr95, upr95), ~ round(., 2)))
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(br_auc$brier_auc_summary, 9))
```

And now look at a plot of the summaries...

```{r, warning=FALSE, comment=NA, message=FALSE, echo=FALSE}
br_auc$brier_auc_hist
```

### Jensen's inequality, random effects, and marginal predictions

When a GLMM includes random effects and a nonlinear link function (e.g., log or logit), back-transforming population-level (marginal) predictions to the response scale introduces a systematic bias. This occurs because the exponential function curves upward  (e.g., a bowl-shaped or exponential growth curve), and averaging on the log scale and then back-transforming always gives a smaller value than back-transforming individual predictions and then averaging. This property is known as Jensen's inequality. For a log-link model with random effects, this means:

$$E[\exp(\beta_0 + u)] = \exp(\beta_0) \cdot \exp(\sigma^2/2)$$

where $\sigma^2$ is the total random effect variance. In other words, $\exp(\beta_0)$ , the typical marginal prediction, underestimates the true arithmetic mean by a factor of $\exp(\sigma^2/2)$.

This bias is directly analogous to the well-known log-normal back-transformation bias. The math is identical; the only difference is which variance components sit on the log scale, *inside* the back-transformation:

|  | Log-normal LM | Log-link GLMM | Log-normal LMM |
|:---|:---|:---|:---|
| Model structure | $\log(Y_i) \sim \text{Normal}(X\beta, \sigma^2_\varepsilon)$ | $\log(\mu_{ij}) = X\beta + u_j$, $u_j \sim \text{Normal}(0, \sigma^2_u)$ | $\log(Y_{ij}) \sim \text{Normal}(X\beta + u_j, \sigma^2_\varepsilon)$ |
| Log-scale variance | Residual $\sigma^2_\varepsilon$ | Random effect $\sigma^2_u$ | **Both:** $\sigma^2_\varepsilon + \sigma^2_u$ |
| Naive back-transform | $\exp(X\beta)$ | $\exp(X\beta)$ | $\exp(X\beta)$ |
| True arithmetic mean | $\exp(X\beta + \sigma^2_\varepsilon/2)$ | $\exp(X\beta + \sigma^2_u/2)$ | $\exp(X\beta + (\sigma^2_\varepsilon + \sigma^2_u)/2)$ |
| Correction factor | $\exp(\sigma^2_\varepsilon/2)$ | $\exp(\sigma^2_u/2)$ | $\exp((\sigma^2_\varepsilon + \sigma^2_u)/2)$ |

In the log-normal LM case, the log-scale variance is the residual variance and the correction is widely known. In the log-link GLMM case, the log-scale variance is the random effect variance and the same correction applies; however, this connection seems to be rarely made explicit in ecological statistics training.

The third column is where the two meet. A log-normal model *with* random effects is subject to both sources of bias simultaneously, because both the residual error and the random effects are additive on the log scale and therefore both sit inside the exponential. The correction is $\exp((\sigma^2_{\varepsilon} + \sigma^2_{u}) / 2)$, and omitting the residual term is a common and consequential mistake: residual variance is typically the *larger* of the two components, so a correction built from `VarCorr()` alone can recover only a fraction of the bias while appearing to have addressed it. A fixed-effects GLM with a log link has no random component on the linear predictor and therefore no Jensen bias at all; $\exp(\beta_0)$ is exactly the arithmetic mean.

This bias grows with random effect variance, is invisible to standard diagnostics (DHARMa, AIC, residual checks), and although it does not affect fixed-effect inference or contrasts, it can substantially affect response-scale predictions used for management.

For models with multiple random effects, the variances are additive on the log scale:

$$\text{correction} = \exp\left(\frac{\sigma^2_1 + \sigma^2_2 + \cdots + \sigma^2_k}{2}\right)$$

**This is not a model deficiency** but rather a structural property of GLMMs. Consistent negative RBIAS in `bias_precision()` output signals that this heterogeneity is being ignored in marginal predictions. For purely inferential purposes (contrasts, p-values, effect directions) this does not matter. For calibrated response-scale predictions (abundance maps, density estimates, management outputs), the correction should be applied.

#### When does Jensen bias matter? Substantive vs nuisance random effects

Whether Jensen bias is consequential for a given analysis depends critically on the nature of the random effects in the model and the purpose of the predictions. Users should consider two broad categories:

**Substantive random effects** represent real, ecologically meaningful group-level variation that is relevant to prediction. For these, Jensen bias matters for response-scale predictions because marginalizing over the RE loses real information about the quantity being predicted:

-   Site random effects when predicting abundance or density at specific locations
-   Spatial random effects when generating abundance maps or occupancy surfaces
-   Individual random effects in mark-recapture models when individual-level predictions are of interest
-   Population random effects in hierarchical models where population identity carries biological meaning

**Nuisance random effects** capture study-specific variation that is not relevant to forward predictions or generalization. For these, marginalizing over the RE is conceptually correct and the Jensen bias is largely irrelevant:

-   Year random effects capturing inter-annual variation during a study period: not meaningful when predicting to future years or simulating counts under specified covariate conditions
-   Observer random effects capturing individual variation in detection: not relevant when predicting to new observers
-   Overdispersion random effects (OLREs): capture residual heterogeneity not attributable to any substantive grouping factor
-   Batch or run random effects in laboratory studies: study-specific, not generalizable

The general principle is: **if the random effect represents variation you would want to condition on in a prediction, the Jensen bias matters. If it represents study-specific noise that should not carry forward, marginalizing over it is correct and the Jensen bias is a non-issue.** Whether to apply the bias correction is therefore a scientific judgment call that depends on the nature of the random effects, the purpose of the predictions, and the magnitude of the bias given the estimated RE variance. `bias_precision()` surfaces the bias and provides tools to quantify and correct it; the decision of whether to act on it rests with the user.

#### Which correction applies? Log *link* vs log *response*

The single most important distinction is whether the log appears as a **link function** or as a **transformation of the response**, because this determines
whether the residual variance sits inside or outside the back-transformation. The family alone does not tell you: a Gaussian model can appear on either side of this line.

| Model | Formula | Residual is additive on... | Correction |
|:---|:---|:---|:---|
| Log-link GLM | `y ~ x, family = poisson` | (no log-scale random component) | none needed |
| Log-link GLMM | `y ~ x + (1|site), family = nbinom2` | response scale | $\exp(\sigma^2_u/2)$ |
| Gaussian log **link** | `y ~ x + (1|site), family = gaussian("log")` | response scale | $\exp(\sigma^2_u/2)$ |
| Log-normal LM | `log(y) ~ x` | **log scale** | $\exp(\sigma^2_\varepsilon/2)$ |
| Log-normal LMM | `log(y) ~ x + (1|site)` | **log scale** | $\exp((\sigma^2_\varepsilon + \sigma^2_u)/2)$ |

`gaussian(link = "log")` and `log(y) ~ x` are easy to conflate, as both involve a Gaussian likelihood and an exponential, but they are different models. With a log link the response is untransformed and the residual error is additive on the *response* scale, outside $\exp()$, so it contributes nothing to the re-transformation bias. With a log-transformed response the residual error is additive on the *log* scale, inside $\exp()$, so it contributes fully.

`GLAMMGoF` resolves this automatically by inspecting the left-hand side of the model formula rather than the family or link, so both cases are handled correctly without user intervention. Only a natural log is supported; `log10()` and `log2()` responses are rejected with an informative error, since each requires a different back-transformation.

#### Diagnosing and correcting Jensen's inequality bias

`bias_precision()` provides a `bias_adjust` argument with two options for diagnosing and correcting Jensen's inequality bias. The `"manual"` correction is available for `glmmTMB` and `lme4` models, and for `lm`/`glm` fits to a log-transformed response. A message is automatically triggered when both in-sample and out-of-sample RBIAS are consistently negative (\< -10%), prompting the user to investigate. The recommended diagnostic workflow is:

**Step 1** - Run with defaults (`bias_adjust = "none"`) and inspect RBIAS. A consistently negative RBIAS in both in-sample and out-of-sample performance suggests Jensen's inequality may be contributing.

```{r, warning=FALSE, comment=NA, eval=FALSE}
# Step 1: default marginal predictions, Jensen signal visible
chk_none <- bias_precision(nReps = 100, testModel = countModel_GLMM,
                           testData = countData, method = "holdout",
                           DHARMaPlot = FALSE, bias_adjust = "none", seed = 123)
chk_none$bias_precision_summary
```

**Step 2** - Re-run with `bias_adjust = "manual"`. If RBIAS moves toward zero, Jensen's inequality is confirmed as the source of the negative bias rather than model misspecification or poor generalization. The analytical correction $\exp(V/2)$ is applied to marginal predictions, where $V$ is the sum of the variance components that are additive on the log scale. For a log-link model this is the total random effect variance from `VarCorr()`. For a model fit to a log-transformed response, the residual variance from `sigma()` is included as well, since it too sits inside the back-transformation.

```{r, warning=FALSE, comment=NA, eval=FALSE}
# Step 2: analytical lognormal correction
chk_manual <- bias_precision(nReps = 100, testModel = countModel_GLMM,
                             testData = countData, method = "holdout",
                             DHARMaPlot = FALSE, bias_adjust = "manual", seed = 123)
chk_manual$bias_precision_summary
```

The figure below shows the effect of applying the Jensen correction across a range of random-effect variances, based on a simulation of intercept-only Poisson GLMMs (100 independent datasets × 100 cross-validation replicates per σ value, N = 20,000). The "none" line traces the underlying Jensen bias as it grows with random-effect variance, reaching roughly −40% at σ = 1 for the two-RE case. The "manual" line shows the same simulations after applying the analytical correction: RBIAS is pulled to near zero across the full σ range whenever random effects are present. For a GLM without random effects (left panel) both options behave identically, since the correction factor is exp(0/2) = 1 and does nothing. Note that a small residual positive RBIAS remains under "manual" at higher σ, growing modestly with random-effect variance. This is a consequence of applying a nonlinear correction to an estimated variance component. Even when σ̂² is unbiased, the convexity of the exponential function means E[exp(σ̂²/2)] > exp(σ²/2), so the plug-in correction factor is on average slightly larger than the true correction factor. The magnitude of this effect grows with σ (because the sampling variance of σ̂² grows with σ), which produces the gentle upward drift visible in the "manual" line. The effect is small (within the confidence ribbons shown) and does not indicate a problem with the correction; users applying the correction to a specific dataset can quantify their own residual by inspecting the corrected RBIAS in bias_precision() output. 

```{r jensen-figure, fig.width = 11, fig.height = 5, echo = FALSE, message = FALSE, warning = FALSE}
# Simulation results generated by data-raw/sim_jensen_sigma_sweep.R

sim_jensen <- readRDS(system.file("extdata", "sim_jensen_sigma_sweep.rds",
                                  package = "GLAMMGoF"))

pal <- c('No correction  ("none")' = "#D85A30",
         'Analytical  ("manual")'  = "#185FA5")

ggplot(sim_jensen$results_df,
       aes(x = sigma_true, y = rbias_mean,
           colour = bias_adjust, fill = bias_adjust)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_fill_manual(values = pal, name = NULL) +
  scale_x_continuous(
    name   = expression(paste("True random-effect ", sigma)),
    breaks = round(sim_jensen$metadata$sigma_grid, 2)
  ) +
  labs(y = "Out-of-sample RBIAS (%)") +
  facet_wrap(~ scenario) +
  theme_bw(base_size = 11) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "grey95"),
        strip.text       = element_text(size = 9))
```

Once Jensen's inequality has been diagnosed as the source of negative RBIAS, the correction can be applied directly to response-scale predictions outside of GLAMMGoF. Note that although the calculations are explicitly written out below, GLAMMGoF has a `jensen_correct()` function which provides a convenient wrapper for this calculation, resolving the appropriate correction automatically from the model and accepting predictions, standard errors, and confidence interval bounds on the link or response scale (see ?jensen_correct for details):

```{r, warning=FALSE, comment=NA, eval=FALSE}
# Manual lognormal correction (fast, reliable for random intercept models)
vc         <- VarCorr(countModel_GLMM)$cond
correction <- exp(sum(sapply(vc, function(x) x[1, 1])) / 2)
preds_adjusted <- predict(countModel_GLMM, newdata = countData,
                          type = "response", re.form = ~0) * correction

# Standard errors for manually corrected predictions should also be scaled
# by the same correction factor, since it is a multiplicative constant.
# The delta method gives: SE_corrected = SE_original * correction
preds_se   <- predict(countModel_GLMM, newdata = countData,
                      type = "response", re.form = ~0, se.fit = TRUE)$se.fit
preds_se_adjusted <- preds_se * correction
```

#### A note about `holdout` vs `bootstrap` resampling methods when diagnosing and adjusting for Jensen's inequality bias

Note on method choice for random-effect models: `method = "holdout"` is recommended over `method = "bootstrap"` when the design has few observations per random-effect group. Case bootstrap with replacement duplicates rows within re-sampled training sets, and these duplicates appear to the model as within-group replicates. When groups are sparse, each duplicated row has disproportionate leverage over that group's apparent within-group structure, systematically inflating the estimated random-effect variance in refitted training models relative to the full-data fit. When bias_adjust = "manual" is used, the correction factor is computed once from the full-data model and therefore under-corrects the inflated per-refit Jensen bias under bootstrap, leaving residual negative RBIAS in the output. The effect is largely independent of model specification; it appears in both correctly-specified and misspecified fits when groups are sparse, and diminishes substantially when groups have many observations regardless of specification. The issue is particularly acute for balanced repeated-measures designs (e.g., a fixed number of sub-samples per site or subject), where case bootstrap generates duplicated rows at rates and structures inconsistent with the sampling design. Cluster bootstrap (resampling entire groups) or parametric bootstrap would be more appropriate resampling schemes; `GLAMMGoF` currently supports neither, so `method = "holdout"` is the recommended choice.

The figure below shows this pattern across a controlled simulation. Panel A shows out-of-sample RBIAS after Jensen correction; Panel B shows per-refit random-effect variance estimates. The bootstrap-vs-holdout divergence appears in both panels when groups are sparse (top row) and largely disappears when groups are dense (bottom row), across both correctly-specified and misspecified fits. 

```{r bootstrap-figure, fig.width = 8, fig.height = 11, echo = FALSE, message = FALSE, warning = FALSE}
# Simulation results generated by data-raw/sim_bootstrap_holdout.R
# See metadata field for seed and session info.

sim_data <- readRDS(system.file("extdata", "sim_bootstrap_holdout.rds",
                                package = "GLAMMGoF"))

# Panel A: RBIAS (manual) by method x family x n_obs
panelA <- sim_data$rbias_df |>
  filter(Group == "Out-of-sample performance") |>
  ggplot(aes(x = method, y = value, fill = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.size = 0.4, alpha = 0.7) +
  facet_grid(n_obs ~ fit_family) +
  labs(x = NULL,
       y = "Out-of-sample RBIAS (%)\nbias_adjust = 'manual'",
       title = "A. RBIAS after Jensen correction, by method") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none")

# Panel B: per-refit RE variance by method x family x n_obs
panelB <- sim_data$revar_df |>
  ggplot(aes(x = method, y = refit_re_var, fill = method)) +
  geom_hline(aes(yintercept = full_data_re_var),
             linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.size = 0.4, alpha = 0.7) +
  facet_grid(n_obs ~ fit_family, scales = "free_y") +
  labs(x = NULL,
       y = "Per-refit RE variance",
       title = "B. Per-refit random-effect variance, by method") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none")

panelA / panelB
```

#### A worked example: correcting for a single random intercept

The only difference here is that now we are setting `bias_adjust = "manual"`

```{r bp-countModelGLMM-corrected}
bp_countModelGLMM_corrected <- bias_precision(
  nReps = 100, 
  testModel = countModel_GLMM, 
  testData = countData, 
  propTrain = 0.8, 
  DHARMaPlot = FALSE, 
  seed = 123, 
  method = "holdout",
  bias_adjust = "manual")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
bp_countModelGLMM_corrected$bias_precision_summary <- bp_countModelGLMM_corrected$bias_precision_summary %>% mutate(across(c(mn, lwr95, upr95), ~ round(., 2)))
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bp_countModelGLMM_corrected$bias_precision_summary, 8))
```

And compare this to results using the default setting of `bias_adjust = "none"`

```{r bp-countModelGLMM-uncorrected}
bp_countModelGLMM_uncorrected <- bias_precision(
  nReps = 100, 
  testModel = countModel_GLMM, 
  testData = countData, 
  propTrain = 0.8, 
  DHARMaPlot = FALSE, 
  seed = 123, 
  method = "holdout",
  bias_adjust = "none")
```

```{r, warning=FALSE, comment=NA, echo=FALSE}
knitr::kable(head(bp_countModelGLMM_uncorrected$bias_precision_summary, 8))
```

RBIAS shifts from roughly −4.5% to +1–2% for both in-sample and out-of-sample performance. (For reference, countModelGLMM has σ̂ ≈ 0.36, giving a correction factor of exp(σ̂²/2) ≈ 1.07. This is a modest ~7% multiplicative shift that is nonetheless enough to move RBIAS from clearly negative to near zero.) The correction has resolved the systematic underprediction from Jensen's inequality; the small residual positive bias reflects the convexity of exp() applied to a plug-in σ̂², discussed further below. Precision metrics also shift with the r-escaling, but it is the change in RBIAS that tells you the Jensen's inequality bias has been resolved.    

#### A worked example: the doubly-biased case

The example below simulates data from a log-normal LMM with known variance components, then compares the naive back-transform against the correction built from the random effect variance alone and the full correction including the residual term.

```{r, warning=FALSE, message=FALSE, comment=NA}
library(glmmTMB)

set.seed(42)
n_site <- 40; n_per <- 25
site   <- factor(rep(seq_len(n_site), each = n_per))
u      <- rnorm(n_site, 0, 0.5)                      # RE sd      = 0.5
x      <- rnorm(n_site * n_per)
logy   <- 2 + 0.3 * x + u[as.integer(site)] +
          rnorm(n_site * n_per, 0, 0.6)              # residual sd = 0.6
lnDat  <- data.frame(y = exp(logy), x = x, site = site)

m_ln <- glmmTMB(log(y) ~ x + (1 | site), family = gaussian, data = lnDat)

s2_u   <- VarCorr(m_ln)$cond$site[1, 1]              # random effect variance
s2_eps <- sigma(m_ln)^2                              # residual variance

naive     <- exp(predict(m_ln, type = "link", re.form = ~0))
re_only   <- naive * exp(s2_u / 2)                   # incomplete correction
corrected <- naive * jensen_correct(m_ln)            # exp((s2_eps + s2_u) / 2)

data.frame(
  quantity = c("Observed mean of y", "Naive back-transform",
               "Corrected: RE variance only", "Corrected: RE + residual"),
  value    = round(c(mean(lnDat$y), mean(naive),
                     mean(re_only), mean(corrected)), 2)
)
```

The naive back-transform recovers the geometric mean and underestimates the arithmetic mean substantially. The RE-only correction closes part of the gap, which is what makes it dangerous: the result moves in the right direction and looks plausible, but a sizable bias remains because the residual variance here is larger than the random effect variance. Only the full correction recovers the observed mean.

`bias_precision()` applies the same correction internally when `bias_adjust = "manual"` is used with a log-transformed response, and reports predictive performance on the original response scale:

```{r, warning=FALSE, comment=NA, eval=FALSE}
bp_ln_none <- bias_precision(nReps = 100, testModel = m_ln, testData = lnDat,
                             method = "holdout", DHARMaPlot = FALSE,
                             bias_adjust = "none", seed = 123)

bp_ln_man  <- bias_precision(nReps = 100, testModel = m_ln, testData = lnDat,
                             method = "holdout", DHARMaPlot = FALSE,
                             bias_adjust = "manual", seed = 123)

bp_ln_none$bias_precision_summary   # RBIAS strongly negative
bp_ln_man$bias_precision_summary    # RBIAS pulled toward zero
```

One limitation applies to log-normal models specifically: `conditional_predictions = TRUE` remains incompatible with `bias_adjust = "manual"`. Note that the reasoning differs slightly from the log-link case: conditional predictions absorb the random effect contribution but *not* the residual contribution, since the residual is always marginalized. A conditional log-normal prediction therefore still requires a residual-only correction of $\exp(\sigma^2_\varepsilon/2)$, which is not currently implemented.

One case cannot be resolved automatically: a response that was log-transformed *before* fitting and stored as its own column. The formula then gives no indication that a transformation occurred, so `jensen_correct()` stops with an error rather than guessing. Pass `type = "lognormal"` explicitly in that situation:

```{r, warning=FALSE, comment=NA, eval=FALSE}
lnDat$log_y   <- log(lnDat$y)
m_precomputed <- lm(log_y ~ x, data = lnDat)

jensen_correct(m_precomputed)                      # error: transformation not visible
jensen_correct(m_precomputed, type = "lognormal")  # correct: exp(sigma^2_resid / 2)
```

#### Potential sources of residual bias after correction

The analytical log-normal correction closes most of the marginal bias in log-link mixed models, but residual negative bias can persist under specific conditions. When you observe uncorrected bias remaining after applying jensen_correct() or the equivalent internal correction, the sources below are the most likely candidates. None of these are problems with your model per se — they are conditions under which the sum-of-variances correction has known limitations, and identifying which one applies tells you something useful about where your response-scale predictions may be unreliable.

A note on how to use this diagnostic: RBIAS and related metrics are informative about where the correction is falling short and, by extension, where response-scale predictions from the model may not be delivering what you expect. That is a legitimate and useful thing to know. What they are not is a model-selection criterion. Iterating on random-effect structure until RBIAS looks better risks removing real structure from the model to optimize a metric, which defeats the purpose. Design-based considerations (does the grouping capture genuine non-independence, do you need to predict to new levels, does partial pooling matter for your inference) and standard fit-based criteria remain the appropriate basis for random-effect specification. Use the diagnostic to understand your model, not to reshape it.

#### Random effects with few levels (< ~10)

The correction depends on σ̂² from the fitted model, and variance components estimated from small numbers of groups are downward-biased and highly variable. REML partially corrects small-sample bias but cannot fully compensate below roughly 10 levels. An under-estimated σ̂² produces under-correction, which shows up as residual negative bias on the response scale.

To investigate: refit the model with the small-group factor as a fixed effect and compare. If fixed-effect estimates remain stable, residual variance does not inflate substantially, and the residual bias resolves, the fixed-effect specification was probably more appropriate on modeling grounds anyway — grouping factors with fewer than ~5–8 levels are commonly recommended as fixed regardless. If fixed effects shift or residual variance inflates, the random effect is capturing structure the fixed-effect specification cannot absorb, and the residual bias is a correction limitation to report alongside your predictions rather than a modeling problem to fix.

#### Random slopes

`jensen_correct()` currently applies a sum-of-variances correction that treats random-slope variance as if it contributed uniformly across observations. The exact marginal correction for a random slope depends on the covariate values at which predictions are made, and intercept–slope covariance complicates it further. This is documented as a known approximation and per-observation slope corrections are planned for a future release. Residual bias in models with random slopes is often traceable to this approximation rather than to any of the other sources listed here.

#### Non-Gaussian random-effect distributions

The analytical correction assumes the random effects are Gaussian. In ecological data, conditional modes are often visibly skewed or heavy-tailed — a few sites, years, or individuals are genuine outliers relative to a Gaussian working assumption. When the true random-effect distribution departs substantially from Gaussian, exp(0.5 Σσ̂²) can under- or over-correct depending on the direction of the departure. A `qqnorm()` check on the conditional modes for each grouping factor is a quick visual diagnostic; substantial non-normality is a signal that the correction is not the whole story and that the model itself may benefit from revisiting (e.g., additional fixed effects that absorb the outlying structure).

#### Unmodeled overdispersion

If Pearson dispersion — sum(resid(model, "pearson")^2) / df.residual(model) — exceeds ~1.2 in a Poisson model, some residual marginal bias on the response scale is coming from un-modeled overdispersion rather than from the correction itself. Refitting as negative binomial and re-running validation typically resolves it.


### Marginal vs conditional predictions: the `conditional_predictions` argument

By default, `bias_precision()` and `brier_auc()` use population-level (marginal) predictions for both in-sample and out-of-sample performance, setting random effects to zero (`re.form = ~0`). This assesses how well the model generalizes to new observations where group-level context is unavailable - the most relevant question when predicting to new sites, new years, or new individuals.

Setting `conditional_predictions = TRUE` switches both in-sample and out-of-sample predictions to condition on estimated random effects (`re.form = NULL`). Since GLAMMGoF's holdout CV splits rows within groups rather than groups themselves, random effect estimates from the training model are valid for held-out observations from the same groups. In-sample and out-of-sample metrics remain directly comparable since both use the same prediction strategy; the gap between them reflects genuine row-level over-fitting rather than any marginal vs conditional distinction.

```{r, warning=FALSE, comment=NA, eval=FALSE}
# Conditional predictions for within-group accuracy assessment
chk_cond <- bias_precision(nReps = 100, testModel = countModel_GLMM,
                           testData = countData, method = "holdout",
                           DHARMaPlot = FALSE, conditional_predictions = TRUE,
                           seed = 123)
chk_cond$bias_precision_summary
```

#### Summary of `bias_adjust` and `conditional_predictions` combinations

The table below summarizes the behavior of each combination of `bias_adjust` and `conditional_predictions` in `bias_precision()`. There are four distinct behaviors in total; the combination marked ❌ is blocked with an informative error.

| `bias_adjust` | `conditional_predictions` | Prediction type | What it measures |
|:-----------------|:-----------------|:-----------------|:-----------------|
| `"none"` (default) | `FALSE` (default) | Marginal | Population-level (i.e., ignore REs) generalization; Jensen signal visible in RBIAS |
| `"manual"` | `FALSE` | Marginal + corrected | Population-level (i.e., ignore REs) generalization with analytical Jensen correction |
| `"none"` | `TRUE` | Conditional | Within-group (i.e., predict to known RE groupings) accuracy; Jensen bias absorbed by REs |
| `"manual"` | `TRUE` | ❌ Error | Double-correction - not permitted |

The `bias_adjust` and `conditional_predictions` restrictions for log-transformed-response models (no `"manual"` with conditional predictions) are described in the worked example above. For `brier_auc()`, the same logic applies, except that `bias_adjust = "manual"` is not available for binomial models regardless of `conditional_predictions`: the analytical correction is only valid where predictions are back-transformed via the exponential (log-link and log-transformed-response models), and the logit link does not admit a closed-form scalar marginal-mean correction. No bias adjustment is currently supported for binomial GLMMs in `brier_auc()`; Brier score and log loss should be interpreted relative to the null model baseline that the function reports alongside the resampling summaries.

The default combination (`bias_adjust = "none"`, `conditional_predictions = FALSE`) is recommended as the starting point for all analyses. It provides the most diagnostic information and surfaces Jensen's inequality bias when present. The other combinations are best used in sequence as part of the step-wise diagnostic workflow described above.

### Recommended workflow for inference and predictions

The following general workflow is recommended for log-link GLMMs with random effects, and applies equally to models fit to a log-transformed response. Use marginal predictions (`re.form = ~0`) for all inferential purposes; fixed-effect contrasts, effect directions, p-values, and model comparison are unaffected by Jensen's inequality bias. For response-scale quantities intended for reporting, figures, or management; predicted abundances, density estimates, and covariate effect plots; apply the log-normal bias correction via `jensen_correct()` or `bias_adjust = "manual"` in `bias_precision()`. This two-stage approach separates the inferential and predictive roles of the model and ensures that response-scale outputs are calibrated to the arithmetic mean of the population rather than the geometric mean.
